---
title: "R Notebook"
output:
  word_document: default
  html_notebook: default
---

*Run*  *Cmd+Shift+Enter*. 
*Insert Chunk*  *Cmd+Option+I*.
*Preview*  *Cmd+Shift+K*

1. Probability that a recombination event occurs between two adjacent base pairs is small, on the order on 10^-8...also approximately 3 billion base pairs where recombination can occur

  A. Assume there are 3B positions where recombination can occurs and the probability for it is 10^-8...recombination events are independent of each other...total # of recombination events during once instance of meiosis is binomially distributed
    
    i. Use dbinom with parameters size = 3000000000, prob = 0.00000001 and plot the PMF for the number of recombination rates per meiosis
```{r}
size <- 3000000000
prob <- 0.00000001
lambda <- 30

x <- seq(0, 60, by=1)
y <- dbinom(x, size, prob)

plot(x, y, type="h", lwd=2, main="Binomial PMF", xlab="Recombination Events", ylab="Probability")
```
    
    ii. Use pbinom with the same parameters to calculate the probability that the number of recombination events per meiosis is 20 or less
```{r}
pbinom(20, size, prob)
```
    
    iii. Use pbinom with the same parameters to calculate the probability that the number of recombination events per meiosis is 40 or more
```{r}
1 - pbinom(39, size, prob)
```
    
    iv. Use qbinom with the same parameters to calculate the 1st percentile
```{r}
qbinom(0.01, size, prob)
```
    
    v. Use qbinom with the same parameters to calculate the 99th percentile
```{r}
qbinom(0.99, size, prob)
```
    
    
  B. Approximate the binomial in part A with the Poisson with parameter ðœ†= 30...30=3 billion x 10^-8
  
    i. Use dpois with parameter lamba = 30 and plot the PMF for the number of recombination events per meiosis
```{r}
y <- dpois(x, lambda)

plot(x, y, type="h", main="Poisson PMF", xlab="Recombination Events", ylab="Probability")
```
    
    ii. Use ppois with the same parameters to calculate the probability that the number of recombination events per meiosis is 20 or less
```{r}
ppois(20, lambda)
```
    
    iii. Use ppois with the same parameters to calculate the probability that the number of recombination events per meiosis is 40 or more
```{r}
1 - ppois(39, lambda)
```
    
    iv. Use qpois with the same parameters to calculate the 1st percentile
```{r}
qpois(0.01, lambda)
```
    
    v. Use qpois with the same parameters to calculate the 99th percentile
```{r}
qpois(0.99, lambda)
```
    
    vi. Does Poisson appear to be a good approximation to the binomial for this problem?
      Yes, the results are identical so the Poisson approximation is valid



2. The exponential density function is (see pdf)...expected value (same as random variable mean) is 1/lambda. Use the dexp function with parameter rate = 1 (corresponds to lambda = 1) to plot the density function. Let n = 100

  A.
    i. Simulate n exponential random variables (use rexp with rate = 1) and take their sample mean...repeat this 1,000 times
```{r}
set.seed(123) # reproducibility
n <- 100
sample_means_100 <- replicate(1000, mean(rexp(n, rate = 1)))
```
  
    ii. Plot the histogram of the 1,000 sample means from i
```{r}
hist(sample_means_100, breaks = 30, main="Sample means (n = 100", xlab="Mean")
```
  
    iii. Calculate the min, mean, an the max of 1,000 sample means from i
```{r}
summary(sample_means_100)
```
  
    iv. Repeat steps i - iii for n = 1000 and n = 10000. Does the sample mean appear to converge?
```{r}
n <- 1000
sample_means_1000 <- replicate(1000, mean(rexp(n, rate = 1)))

hist(sample_means_1000, breaks = 30, main="Sample Means (n = 1000)", xlab="Mean")
summary(sample_means_1000)
```
```{r}
n <- 10000
sample_means_10000 <- replicate(1000, mean(rexp(n, rate = 1)))

hist(sample_means_10000, breaks = 30, main="Sample Means (n = 10000)", xlab="Mean")
summary(sample_means_10000)
```
      The sample mean stabilizes around 0.97 - 1.0 as n increases and the spread of the histogram decreases.

  
  
  B. The pareto distribution is (see pdf)...isn't the standard in R...need to install.packages("EnvStats)...when a = 1 the expected value is undefined (or infinity)...recall the Law of Large Numbers and the Central Limit Theorem
  
    i. Simulate n Pareto random variables (use rpareto with location = 1, shape = 1) and take their sample mean. Repeat this 1000 times
```{r}
#install.packages("EnvStats") # hash after running once
library(EnvStats)

set.seed(123) # for reproducibility
n <- 100
sm_pareto_100 <- replicate(1000, mean(rpareto(n, location = 1, shape = 1)))
```
    
    ii. Plot the histogram of the 1000 sample means from i
```{r}
hist(log10(sm_pareto_100), breaks=10, main="Pareto Sample Mean (n = 100)", xlab="Mean")
```
    
    iii. Calculate the min, mean, and the max of 1,000 sample means from i
```{r}
summary(sm_pareto_100)
```
    
    iv. Repeat steps i - iii for n = 1000 and n = 10000. Does the sample mean appear to converge?
```{r}
n <- 1000
sm_pareto_1000 <- replicate(1000, mean(rpareto(n, location = 1, shape = 1)))

hist(log10(sm_pareto_1000), breaks = 50, main="Pareto Sample Means (n = 1000)", xlab="Means", probability=TRUE)
summary(sm_pareto_1000)
```
```{r}
n <- 10000
sm_pareto_10000 <- replicate(1000, mean(rpareto(n, location = 1, shape = 1)))

hist(log10(sm_pareto_10000), breaks = 50, main="Pareto Sample Means (n = 10000)", xlab="Means", probability=TRUE)
summary(sm_pareto_10000)
```
    
    
    
    
3. Consider the simple model for father and son heights. Let X be a normal random variable with mean 70 and sdev o, let e1 be a normal random variable with mean 0 and sdev t, and let e2 be a normal random variable with mean 0 and sdev t. Further assume X, e1, and e2 are independent. Define F = X + e1 and S = X + e2...idea is that if F and S are father and son heights, respectively, they both have the same X due to genetics and independent 'disturbance' terms e1 and e2

  A. Calculate the covariance of F and S
```{r}
set.seed(123)
n <- 10000

sigma <- 3 #sdev of genetic component
tau <- 2 #sdev of environmental variation

X <- rnorm(n, mean = 70, sd = sigma) # Genetic
epsilon_F <- rnorm(n, mean = 0, sd = tau) # Random effect father
F <- X + epsilon_F # father's height
epsilon_S <- rnorm(n, mean = 0, sd = tau) # Random effect son
S <- X + epsilon_S # son's height

# Covariance calculation
cov(F, S) # should be close to sigma^2
```
  
  B. Calculate the correlation of F and S
```{r}
cor(F, S) # Should be <1 due to regression

expected_corr <- sigma^2 / (sigma^2 + tau^2)
expected_corr
```